---
author:
- Jon Zelner
header-includes:
- \usepackage{float}
- \usepackage{wrapfig}
- \usepackage{graphicx}
- \usepackage{appendixnumberbeamer}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty, font=footnotesize}
title: The three paths of hierarchical modeling
institute: EPID 684 \newline University of Michigan School of Public Health  \newline \newline `jzelner@umich.edu` \newline `www.jonzelner.net`
fontsize: 12pt 
output: 
 beamer_presentation:
    latex_engine: xelatex
theme: metropolis
date: \today
---

## Agenda

- Hierarchical GLM Notation

\pause

- Likelihood and model fit activity

\pause

- Radon! 

# A re-introduction to Generalized Linear Models (GLMs) for *clustered* data 

## Notation for a classic un-clustered GLM

Going to be seeing a lot of this:

- $y_{i} = \alpha + \beta x_i + \epsilon_i$


\pause

Where:

 - $y_i$ is \alert{continuous} outcome measure: height, BMI, etc.
 \pause
 - $\beta$ is risk associated with some kind of exposure 
 \pause
 - $x_i \in [0,1]$ is an indicator of exposure.
 \pause
 - $\alpha$ is expected outcome when $x_i$ = 0
 \pause
 - $\epsilon_i$ are independently and identically distributed (i.i.d.) errors

## Independent errors

Classic assumption is that:

- $\epsilon_i \sim N(0, \sigma^2)$

\pause

In plain-ish English:

 - Observation $y_{ij}$ of individual $i$ is a function of $\alpha + \beta x_i$ and normally distributed \alert{errors} ($\epsilon_i$) with mean zero and variance $\sigma^2$.

\pause

Another way of writing it:

- $y_i \sim N(\alpha + \beta x_i, \sigma^2)$

## Three Approaches to Modeling Clustered Data

\begin{figure}
  \captionsetup{justification=justified}
  \includegraphics[height=0.5\textwidth]{images/letsmakeadeal.jpg}
  \caption{Which door will you choose?}
\end{figure}


## Door #1: Ignore clustering and fit a normal GLM

- *Pool* data across all units, i.e. ignore clustering.

- i.e. fit model $y_{ij} = \alpha + \beta x_i + \epsilon_i$

\begin{center}
Is this a good idea? Why or why not?
\end{center}

## NO!

\begin{figure}
  \captionsetup{justification=justified}
  \includegraphics[height=0.6\textwidth]{images/regression_dragon_1.jpg}
  \caption{Complete pooling ignores potential sources of \textit{observed} and \textit{unobserved.} unit-level \alert{confounding}.}
\end{figure}

## Pooling clustered data violates assumption of independent errors

A \alert{pooled} model:

\begin{equation}
y_i = \alpha + \beta x + \epsilon_i
\end{equation}

\pause

- $y_i$ is a combination of systematic variation ($\alpha + \beta x$) and *uncorrelated* random noise ($\epsilon_i$) where:


\begin{equation}
i.i.d.\ \epsilon \sim Normal(0, \sigma^2)
\end{equation}

## Clustering may result in correlation between average differences from mean

\begin{figure}
  \includegraphics[height=0.6\textwidth]{images/merlo_variance_partition.png}
\end{figure}


## Your \alert{residuals} should look like this

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Example of residuals for model with clustered errors"}
require(ggplot2)
df <- data.frame(x = rnorm(1000))
g <- ggplot(df, aes(x = x)) +
    geom_histogram(binwidth = 0.1) +
    xlab("Distance from mean model prediction") +
    ylab("N")

plot(g)
```

## When you ignore clustering you may see something like:

```{r, echo=FALSE, fig.cap="Clustered errors"}
require(ggplot2)
ind_cluster <- 100
ncluster <- 10
cluster_sigma <- 1
ind_sigma <- 0.2

cluster_ids <- sort(rep(1:ncluster, ind_cluster))

cluster_means <- rnorm(ncluster, sd = cluster_sigma)
ind_vals <- rnorm(n = length(cluster_ids), mean = cluster_means[cluster_ids], sd = ind_sigma)

df <- data.frame(x = ind_vals, cluster = cluster_ids)

g <- ggplot(df, aes(x = x, cluster = cluster_ids)) +
    geom_histogram(binwidth = 0.05) +
    xlab("Distance from mean") +
    ylab("N")

plot(g)
```

## Door #2: Fit a different model to each cluster

Fit *unpooled* model to each unit ($j$), assuming outcomes in each unit are independent:

- $y_{ij} = \alpha_j + \beta_j x_i + \epsilon_{ij}$

- $\epsilon_{ij} \sim N(0, \sigma_{j}^2)$

## More danger!

\begin{figure}
  \captionsetup{justification=justified}
  \includegraphics[height=0.6\textwidth]{images/regression_dino.jpg}
  \caption{Totally unpooled models run the risk of \alert{overfitting} the data, particularly in small samples.}
\end{figure}

## Specific dangers of unpooled models

What else could go wrong here?

\pause

- Some units (e.g. counties) may have few observations, making *unpooled* models impractical

\pause

- We may want to allow some effect of exposure (e.g. having a basement) to be consistent across counties. 

## Door #3: Partial Pooling!

- Allow effects to vary across clusters, but constrain them with a \alert{prior} distribution.

\pause

- This approach accommodates variation across units without assuming they have no similarity.

\pause

- More likely to make accurate \alert{out-of-sample} predictions than the fully-pooled or unpooled examples.

## Partial pooling = \alert{Regularization}

\begin{figure}
  \captionsetup{justification=justified}
  \includegraphics[height=0.6\textwidth]{images/wiki_regularization.png}
  \caption{Both functions fit the data perfectly...which one should you prefer?}
\end{figure}

# Radon Example

## Radon is a carcinogenic gas 
\begin{figure}
  \captionsetup{justification=justified}
  \includegraphics[height=0.5\textwidth]{images/radon_entry.png}
  \caption{Radon is a byproduct of decaying soil uranium.}
\end{figure}

## Radon enters a house more easily when it is built into the ground
\begin{figure}
  \captionsetup{justification=justified}
  \includegraphics[height=0.6\textwidth]{images/basement_radon.png}
  \caption{Ann Arbor is a radon hotspot!}
\end{figure}

## Considerable geographic variation in radon potential
\begin{figure}
  \captionsetup{justification=justified}
  \includegraphics[height=0.6\textwidth]{images/radon_map.png}
  \caption{Ann Arbor is a radon hotspot!}
\end{figure}

## Trust me on this one...
\begin{figure}
  \captionsetup{justification=justified}
  \includegraphics[height=0.6\textwidth]{images/jz_radon.jpg}
  \caption{My very own radon mitigation system.}
\end{figure}


## What should a model that accounts for important sources of variation in household radon potential include?

\pause

- County-level variation in soil uranium.

- Whether or not the radon measurement was taken in a basement.


## \alert{Random intercepts} account for county-level variation

Gelman [@Gelman2006] proposes a multi-level model to measure household radon in household $i$ in county $j$, $y_{ij}$:

- $y_{ij} \sim N(\alpha_j + \beta x_{ij}, \sigma^{2}_y)$, for $i = 1, \ldots, n_j$, $j=1,\ldots,J$
\pause

Where:

- $\alpha_j$ is average, \alert{non-basement} radon measure at county level

- $\beta$ is \alert{fixed effect} measuring average change in radon level in houses with a basement.

- $\sigma^{2}_y$ represents \alert{within-county} variation in risk


## Include predictors of county-level variation in second level

\alert{County-level random intercept} is a function of county soil uranium measure, $u_j$:

- $\alpha_j \sim N(\gamma_0 + \gamma_1 u_j, \sigma_{\alpha}^2)$, for $j = 1, \ldots, J$

\pause

Where:

- $\gamma_0$ is expected \alert{household radon measure} when $u_j = 0$
- $\gamma_1$ scales expected county-level uranium with $u_j$
- $\sigma_{\alpha}^2$ is \alert{between-county} variation in radon risk not measured by $u_j$.

## Putting it all together

\alert{County-level intercept} is a function of county soil uranium measure, $u_j$:

- $\alpha_j \sim N(\gamma_0 + \gamma_1 u_j, \sigma_{\alpha}^2)$

\pause

\alert{Household-level} radon measure is a function of having a \alert{basement} and county-level intercept: 

- $y_{ij} \sim N(\alpha_j + \beta x_{ij}, \sigma^{2}_y)$

## County-level radon levels vary with soil uranium measures

\begin{figure}
  \captionsetup{justification=justified}
  \includegraphics[height=0.5\textwidth]{images/gelman_county_uranium.png}
  \caption{County-level intercept, $\alpha_j$, ($\pm 1$ standard error) as a function of county-level uranium.}
\end{figure}

## Model predictions vs. radon measures by county

\begin{figure}
  \captionsetup{justification=justified}
  \includegraphics[height=0.4\textwidth]{images/gelman_basement_effect.png}
  \caption{Multi-level regression line, $y = \alpha_j + \beta x$, from 8 Minnesota counties. Unpooled estimates = light grey line; Totally pooled estimates = dashed grey line.}
\end{figure}

## Next Time

- Hands-on with the Radon example

# References

## References {.allowframebreaks}

