---
title: "Understanding the role of simulation in statistical inference."
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Sampling from a Normal distribution

In this example, we're going to go back to basics, and use both a formula and simulation to calculate confidence intervals for a *sample mean*. So, first, pick a mean and standard deviation and number of samples to draw from a Normal distribution. 

Move the sliders around to get a sense of how much things change as you incrase the sample size.



```{r, echo=FALSE}

numericInput("mean", "Mean:", value = 0)

numericInput("sd", "Standard Deviation:", min = 0.1, max = 4, value = 1)

sliderInput("nsamp", "Number of samples:", min = 1, max = 500, value = 20)


plotOutput("distPlot")
```

So, for this one sample, we now have  the following sample mean and sample standard deviation:
```{r, echo=FALSE}
textOutput("samplemean")
textOutput("se")
```

```{r, context="server"}

x <- reactive({rnorm(input$nsamp, input$mean, input$sd)})

output$distPlot <- renderPlot({

  hist(x(),  col = 'darkgray', border = 'white', xlab = mean(x))
})

output$samplemean <- renderText({paste0("Sample mean = ", round(mean(x()),2))})

output$se <- renderText({paste0("Sample standard deviation = ", round(sd(x()),2))})

```

### Calculating Confidence Intervals

The classic approach to calculating 95% confidence intervals for a population mean is to look up the appropriate value in a *t-table*, as a function of the number of *degrees of freedom*, i.e. the number of overall samples contributing to the mean.

In R, we can do this as follows, where $s$ is the sample standard deviation and $a$ is the sample mean.

```{r eval=FALSE, echo = TRUE}

## First calculate the standard error using the quantiles of 
## the t-distribution


error <- qt(0.975, df = n-1)*(s/sqrt(n))

## Then calculate upper and lower confidence limits
lower_ci <- a - error
upper_ci <- a + error

```

Now, go ahead and calculate these for the input values above:

```{r}
actionButton("ci", "Calculate Confidence Intervals")
textOutput("sse")
```

```{r, context="server"}
sse <- eventReactive(input$ci, {
  qt(0.975, df = input$nsamp-1)*(sd(x())/sqrt(input$nsamp))
})

output$sse <- renderText({
  paste0("Sample Mean = ", round(mean(x()),2), " Standard error = ", round(sse(), 2)," 95% CI  =", round(mean(x()) - sse(),2), ", ",round(mean(x())+sse(),2))
  }
  )
```

### Yeah, but what does that mean?

We are supposed to interpret these confidence intervals as meaning "95% of the time, the *true mean* will fall between the confidence intervals. So, why don't we test the proposition?

First, let's take a look at the distribution of *sample means* we get from taking 1000 samples:

```{r}
actionButton("sm", "Simulate sample means")

plotOutput("samplemeans")
```

```{r, context = "server"}

sm_data <- eventReactive(input$sm, {
  mean_output <- rep(0, 1000)
  for (i in 1:1000) {
    mean_output[i] <- mean(rnorm(input$nsamp, mean = input$mean, sd = input$sd))
  }
  
  return(mean_output)

})

output$samplemeans <- renderPlot({hist(sm_data(), xlab = paste0("Standard error of the mean = ", round(1.67*sd(sm_data()),2)))})

```

So, if for each of those sample means, we computed the confidence intervals, if all goes to plan, 95% of the time, they should include the true value of the mean you set all the way at the top. We can do this easily using the `TInterval` function from the R `sigr` package, like so:

```{r, eval=FALSE, echo=TRUE}
for (i in 1:1000) {
    sampled_vals <-rnorm(input$nsamp, mean = input$mean, sd = input$sd)
    tvals <- sigr::TInterval(sampled_vals)
    
    if ((input$mean >= tvals$interval_low) && (input$mean <= tvals$interval_high)) {
      num_containing <- num_containing + 1
    }
    
}

```

```{r}
actionButton("simint", "Simulate Confidence Intervals")
textOutput("cipct")
```

```{r, context = "server"}

ci_contains <- eventReactive(input$simint,{
  
  num_containing <- 0
  
  for (i in 1:1000) {
    sampled_vals <-rnorm(input$nsamp, mean = input$mean, sd = input$sd)
    tvals <- sigr::TInterval(sampled_vals)
    
    if ((input$mean >= tvals$interval_low) && (input$mean <= tvals$interval_high)) {
      num_containing <- num_containing + 1
    }
    
  }
  
  return(paste0(round(100*num_containing/1000,2), "%"))
  
})


output$cipct <- renderText({
  paste0("Contains the true mean ", ci_contains(), " of the time!")
  })

```

