---
title: "Simulating Segregation"
output: html_notebook
---

Let's start with a simple scenario in which we have city with 100000 inhabitants, distributed across 400 grid cells representing city blocks on a 24x24 square grid.

Let's start by assigning each grid cell to a neighborhood.

```{r setup}
dissimilarity <- function(x,n) {
  total_a <- sum(x)
  total_b <- sum(n-x)
  d <- 0.5*sum(abs((x/total_a) - ((n-x)/total_b)))
  return(d)
}


half_grid <- function(d, ratio, N, p) {
  
  ## Sample number of unvaccinated in total population
  
  B <- rbinom(1, N, p)
  
  A <- N-B
  
  neighborhood_mat <- matrix(1, d, d)
  ncell <- length(neighborhood_mat)
  neighborhood_mat[, 1:floor(d / 2)] <- ratio
  
  neighborhood_df <- expand.grid(x = 1:d, y = 1:d)
  cell_rates <-
    as.vector(neighborhood_mat[cbind(neighborhood_df$y, neighborhood_df$x)])
  neighborhood_df$b <- rmultinom(1, B, cell_rates)
  neighborhood_df$a <- rmultinom(1, A, 1 / cell_rates)
  
  neighborhood_df$n <- neighborhood_df$a + neighborhood_df$b
  neighborhood_df$group <- neighborhood_df$b / neighborhood_df$n
  
  return(neighborhood_df)
  
}

grid_plot <- function(df) {
  
 g <- ggplot(df, aes(x = x, y = y, fill = group)) + 
  geom_raster() + 
  theme_bw() + 
  coord_equal() +
  guides(fill = guide_legend(title = "Proportion Unvaccinated")) 
 
 return(g)
}

proportion_hist <- function(df) {
  g <- ggplot(df, aes(x=group)) +
    geom_histogram(binwidth=0.005) +
    xlab("Proportion Unvaccinated") +
    ylab("N")
  
  return(g)
}

## Define a function to calculate Moran's I for an 
## arbitrary vector of values on a square grid

morans_grid <- function(d, x) {
  require(spdep)
  
  neighbors <- nb2listw(cell2nb(d,d))
  
  return(moran(x, neighbors, 1, 1))

}

isolation <- function(x,n) {
  x <- x[n > 0]
  n <- n[n > 0]
  total_a <- sum(x)
  d <- sum((x/total_a)*(x/n))
  return(d)
}
```


```{r makegrid}
require(dplyr)
require(ggplot2)
require(spdep)

## Dimension of neighborhood grid
d <- 8

## Size of the total population
total_pop <- 100000
unvaccinated_proportion <- 0.05
ratio <- 1
## Number of vaccinated and unvaccinated individuals


## Generate grid
neighborhood_df <- half_grid(d, ratio, total_pop, unvaccinated_proportion)

g <- grid_plot(neighborhood_df)
plot(g)
```


# What does this look like at the population level?

```{r}
 plot(proportion_hist(neighborhood_df))
```

## How does this translate into *isolation*?
```{r}

iso_val <- isolation(neighborhood_df$b, neighborhood_df$n)

dissim_val <- dissimilarity(neighborhood_df$b, neighborhood_df$n)

print(paste0("Isolation = ", round(iso_val,2)))
print(paste0("Dissimilarity = ", round(dissim_val,2)))
```

## How do these change with the proportion unvaccinated?

```{r}

## Make a vector with a range of unvaccinated fractions from 1% to 40%
unvax_p <- seq(from = 0.01, to = 0.4, by = 0.01)
ratios <- seq(from = 1, to = 5, by = 0.5)
## Empty vectors to store the estimated dissimilary and 
## isolation values
iso_vals <- rep(0, length(unvax_p))
dissim_vals <- rep(0, length(unvax_p))
moran_vals <- rep(0, length(unvax_p))
for (i in 1:length(unvax_p)) {
  
  ## Generate a new grid with the specified unvaccinated fraction
  neighborhood_tmp <- half_grid(d, 3, total_pop, unvax_p[i])
  iso_vals[i] <- isolation(neighborhood_tmp$b, neighborhood_tmp$n)
  dissim_vals[i] <- dissimilarity(neighborhood_tmp$b, neighborhood_tmp$n)
  moran_vals[i] <- morans_grid(d, neighborhood_tmp$group)$I
}

```

## Visualize relationship between unvaccinated proportion and isolation
```{r}
plot(unvax_p, iso_vals)
```

## Dissimilarity

```{r}
plot(unvax_p, dissim_vals)
```

## Moran's I

```{r}
plot(unvax_p, moran_vals)
```

